Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-10-26 src/tex/ms.tex
[showyourwork] overleaf sync

\newcommand{\vlk}[1]{{\color{blue} [VLK: #1]}}

% Begin!
\begin{document}

@@

% Author list
    \author[0000-0003-4568-7005]{Axel Donath}
    \author[0000-0002-0905-7375]{Aneta Siemiginowska}
    \author[0000-0002-3869-7996]{Vinay Kashyap} 
    \author[0000-0000-0000-0000]{David van Dyk}

@@

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is often called \enquote{deblurring} or \enquote{deconvolution}, as it aims to invert the effect of the data being degraded by a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is categorised as \enquote{inverse problems}. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as \enquote{unblind deconvolution}. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant activate galactic nuclei or a model of the PSF can obtained from simulation of the telescope or instrument.

    \subsection{Related Work}
    In literature there have been many efforts to solve the unblind deconvolution problem for images affected by Poisson noise. Most methods rely fundamentally on modeling the image generating process as a linear process. Given a known PSF a prediction of the discrete measured image can be obtained via \enquote{forward folding} and be compared against an observation assuming Poisson statistics per pixel. However the methods differ in their additional prior or regularisation assumptions as well as their approach to optimization or sampling of the underlying likelihood function. In the following section we give a short overview over those methods.

@@

    However in practice the RL method was found to have multiple, partly related limitations: first there is no objective criterion for when to stop the optimization process. Usually the iteration is stopped when the reconstructed image seems plausible to the eye. Second, when the number of iterations is too high, the method shows a tendency to decompose regions of constant brightness into a set of bright emission peaks (\enquote{speckles}). This problem was for example described and studied by \cite{Reeves1995} and \cite{Fish1995}. Lastly the RL method only provide a \enquote{point estimate} and no associated estimate of the uncertainties of the reconstruction, which is desirable especially in the Poisson domain for example to measure the significance of specific features found in the reconstructed image.

    To address the issue with RL decomposing extended emission structures into groups of point source, multiple additions to the RL method were proposed. Most importantly this lead to the introduction of regularised RL methods, such as use of total variation \citep{Dey2006} or sparse representations based on \enquote{wavelets} or \enquote{curvelets} \citep{Starck2003}. The challenge is rooted in two fundamentally contradicting goals: a deconvolution process that favors point sources will decompose extend structures present in the image, while introducing local correlations between neighbouring pixels will lead to less sharp reconstruction of point sources.

    One possible approach to overcome this challenge has been proposed by \cite{Esch2004} and \cite{Connors2011} as the \enquote{Low Counts Image Reconstrction and Analysis} (LIRA) method. Their key idea was to introduce a prior assumption that favors smoothness on multiple spatial scales of the image. Instead of a trying to find a single point estimate for the reconstructed image  they also changed to sampling from the posterior distribution, which allowed them to quantify the associated uncertainties. However the prior distribution is based on the generic assumption of smoothness and does not encode any \enquote{real world} or physical information on the structures present in the image.

    An interesting modification of the standard RL method has been proposed by \cite{Ingaramo2014}. They showed that RL can be effective in creating a merged image from multiple observations. They also observed that RL would take the best visual properties out of an each image and merged them to create a more accurate estimate of the ground truth.

    \cite{Selig2015} and \cite{Pumpe2018} approached the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they were able to deconvolve an astrophysical observation as well as decomposing it into point-like and diffuse emission components. While the method has worked well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it seemed to fail for smaller extended emission structures such as mrophological details of extended sources.

    Other interesting approaches to the problem of deconvolution can of course be found in the field of computer vision. For example \cite{Zoran2011} build on very similar assumption of local correlation for natural images. However instead of assuming a fixed local correlation structure, such as smoothness defined by a Gaussian with a correlation size, they proposed to learn the local correlation structure on millions of patches extracted from a set of reference images. For this they proposed to split training images from representative data sets into small patches and learn a  Gaussian Mixture Model (GMM) modeling the distribution of extracted patches. To use the GMM in reconstruction they introduced the concept of expected patch likelihood (EPLL). They showed that the patch based GMM prior led to much improved image reconstructions for multiple inverse problems such as denoising, deblurring and inpainting. \cite{Bouman2016} later adapted the patch prior successfully for the reconstruction of radio data from the Event Horizon Telescope (EHT).

@@

    \subsection{Some Remarks on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the standard for deblurring and super-resolution of natural images. However their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical x-ray and \gammaray data.

    Firstly, astronomical x-ray and \gammaray data data is rare, expensive to obtain and there is no ground truth, because we can only observe the sky with our given instruments. To circumvent this one could rely on transfer learning and simulated data. For example radio or optical images could be forward folded with a known PSF model and be degraded with Poisson noise to generate training data. However astronomical images still have a large dynamic range and show a high diversity in morphological structure, which would still require a large amount of training data. However this approach has been successfully applied for specific classes of sources, such as images of Galaxies in the Sloan Digital Sky Survey (SDSS) \citep{Schawinski2017} or as a post processing step of results obtained with regularised RL \citep{Akhaury2022}. \cite{Sweere2022} used images from
    the large scale structure simulations with promising results.

@@

    One last important challenge for deep learning based methods is obtaining uncertainty estimates for the reconstructed image from CNNs. In the limit of Poisson noise it is desirable to trace the full a-postiori likelihood, which is not possible for most network architectures, with the exception of for example normalising flows. This problem could be addressed with Monte Carlo (MC) based methods, such as bootstrapping, which further increases the demand on training data and computational resources. For these reasons we decided to choose an approach more close to \enquote{classical} statistical modelling techniques. However we will include elements from ML to represent the prior information on images.

@@

    \begin{itemize}
        \item Different observations of one instrument or telescope at different times and observation conditions. For example, multiple observations of \chandra of the same astrophysical object with different offset angles and exposure times
        \item Observations from different telescopes, which operate in the same wavelength range. For example, a \chandra and \xmm observation of the same region in the sky
        \item A single observation of one telescope with different data quality categories and different associated instrument response functions, such as event classes for \fermi
    \end{itemize}

    Or even an arbitrary combination of the possibilities listed above. It could also be the combination of observations in different energy ranges, however in this case the assumption of the morphology not changing with energy would apply, which is rarely fulfilled for the astronomical data we are dealing with.

@@

    By taking the logarithm and dropping the constant terms one can transform the
    product into a sum over pixels, which is also often called the \textit{Cash}
    \citep{Cash1979} fit statistics:
    %
    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{D_j} | \boldsymbol{\lambda_j} \right) = \frac{1}{N}\sum_{i=1}^N \lambda_{j,i} - D_{j, i} \log{\lambda_{j,i}}
    \end{equation}

    To account for the possibility of different datasets having a different number of pixels we divide by $N$. By summing over all observations we finally get the joint log-likelihood of measuring a set of counts images $\mathbf{D}$ given an unknown flux image $\mathbf{x}$:
    %
    \begin{equation}
        \label{eq:joint}
        \mathcal{L}\left( \mathbf{D} | \mathbf{x} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D_j} | \mathbf{x} \right)
    \end{equation}

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate of $\hat{\mathbf{x}}$ directly, using some optimization procedure. However in general this represents an \textit{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the existence of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

@@

    \subsection{Patch Based Image Prior}
    In general it is difficult to capture global image statistics in a prior likelihood taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the number of pixels squared and thus quickly becomes computationally untraceable. However astronomical images typically only show local correlations between pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light (structure in the CMB might be an exception here). This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the idea of a patch based image prior approach as introduced by \cite{Zoran2011}. Given a pre-trained Gaussian Mixture Model (GMM) for the distribution of pixel values in patches, we can evaluate a patch based prior for the reconstructed image $\mathbf{x}$ and approximate the \aposteriori likelihood.

    One convenient property of the GMM is the possibility to evaluate its likelihood in closed form. For a given single patch $\mathbf{x}_n$ it is given by:
    %
    \begin{equation}
        \label{eq:gmm}
        P_{GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k \mathcal{N}(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2)
    \end{equation}

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the most likely GMM component per patch, becomes a discrete hyper-parameter in the model that is optimized along with the image $\mathbf{x}$. The bottom row of Figure~\ref{fig:patches} shows an example of a patch and the same patch represented in the \enquote{Eigenbasis} of the covariance matrices of its three most likely GMM components. Visually the representation of the of the patch in the Eigenbasis of the covariance results in a denoised version of the patch.

@@

    \cite{Zoran2011} found that the reconstruction quality could be improved by working with a grid of overlapping patches. Such a grid is illustrated in Figure~\ref{fig:patches}. However using a grid of overlapping patches makes the interpretation of the prior as likelihood more difficult, because pixels at the corners and edges of patches are counted multiple times. For this we introduce a weighting mask when evaluating the likelihood of the GMM. This mask weights the pixels of the patch by their overlap with the neighboring patch. Overlapping edges are assigned a weight of 1/2, while overlapping corners are assigned a weight of 1/4, when computing the likelihood for a patch as given by  Equation~\ref{eq:gmm}. Apart from this we keep the process of choosing the most likely component unmodified. However we note that in this case we neglected the correlation between neighbouring patches. As the choice of the GMM component per patch is discrete, a simultaneous gradient based optimization of all patches is not easily possible, because we cannot compute gradients across GMM components. This would require a continuous representation or interpolation between GMM components.

@@

    To normalise the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to present at many more levels of intensity. While this could be handled by introducing a lot more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we propose the following modified normalisation for the patches instead:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \tilde{\mathbf{x}}_n = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}
    \end{equation}

@@

    \subsubsection{Cycle Spinning}
    The evaluation of the patch prior divides the reconstructed image into a fixed grid of overlapping patches. To avoid artifacts in the reconstruction due to the fixed position of the grid, we use the method of \enquote{cycle spinning}. Before evaluating the patch prior we randomly shift the image $\mathbf{x}$ by a given number of pixels in x and y direction. Were the shift is smaller than half the size of a patch. The same method has, for example, been used in \cite{Esch2004}. We also experimented with cycle spinning by a sub-pixel amount, mostly for the reconstruction of point sources. However we later introduced the possibility to reconstruct an upsampled version of $\mathbf{x}$, which made the sub-pixel cycle-spinning mostly redundant.

@@

    \subsection{Systematic Errors of Predicted Counts}
    Each measured counts image $\mathbf{D_j}$ is associated with systematic errors. This includes for example the pointing accuracy of the telescopes: when repeatedly pointed to the same object or when using different types of telescopes for observation, the absolute position of an astronomical object might vary in the image between different observations. The process of correcting for these inaccuracies is called \textit{astrometry} and a standard procedure in for example \chandra data analysis. A second example of systematic errors is the variation of background emission between multiple observations.

@@

    \subsection{Estimation of Statistical Errors}
    Estimating the global covariance matrix for image based data is computationally not feasible, because the size of the matrix scales with the squared number of pixels in the image. However there are a few ways to estimate the statistical uncertainties associated with the flux image reconstructed by \jolideco. We propose three methods for this:

    \begin{itemize}
    \item The first method we propose is bootstrapping of observations. Given that multiple observations of the same region of the sky are available, \jolideco can be run repeatedly on a random subset of those. This results in a set of reconstructed images which are sampled from the posterior distribution. Depending on the size of the images the associated computational effort can be very large as it requires running \jolideco many times to achieve sufficient statistics.

    \item If only one or few observations are available we propose bootstrapping of the events used to compute the counts image instead. This basically corresponds to splitting a single observation into multiple ones splitting the single exposure into multiple time intervals. This way one can apply the first proposed method again.

@@

@@

    \subsection{Jolideco GMM Library}
    \label{ssec:jolideco-gmm-library}
    In general the strength of the GMM prior is its high \textit{expressiveness}. This means the prior contains a large variety of structural shapes and has a lot of freedom to adapt to the data in the optimization process. On radio data \cite{Bouman2016} found their results to be mostly independent of the choice of data the GMM prior was learned on. However in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of the data the GMM was learned on. The choice of the prior becomes more important and to avoid introducing a strong bias it should be adapted to the analysis scenario. For this reason and for purposes of testing we learned and provide a selection of GMMs (a \textit{Library}) to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

@@

    \subsubsection{GLEAM Radio Data}
    \label{sssec:gleam-radio-data}
    As an alternative we consider a prior learned from astronomical radio data. For this we relied on high signal to noise radio data from the \textit{GLEAM} survey \cite{HurleyWalker2022}. We downloaded the data using the Gleam VO Client\footnote{\url{https://github.com/ICRAR/gleamvo-client}}. We restricted the choice of data to the Galactic Plane between $b=\pm10\deg$ and $\ell = \pm180\deg$ to avoid a bias towards extra-galactic point sources and represent structures of Galactic sources with rich morphology, such as supernova remnants, pulsar winds etc. The dataset still includes Galactic as well as extra-galactic point sources in the background. In total we extracted $\approx 6 \cdot 10^5$ patches and trained a GMM with 128 components. We used the patch normalisation described in Section~\ref{ssec:patch-norm}.

@@

    \subsubsection{Chandra SNRs}
    \label{sssec:chandra-snrs}. Because of the limited statistics we did not use any oversampling but kept the pixel Astrophysical processes forming the morphological structures are typically self-similar. Structures appear on multiple angular scales. This means the prior could even be learned from the data itself if sufficiently free of noise. For analysis of Galactic sources with complex morphology and typically radially symmetric features such as supernova remnants (SNRs) we learned a prior from images of the bright \chandra SNRs \textit{Cas A} and \textit{G292.0+01.8}. In total we extracted $\approx 10^6$ patches and learned a GMM with 128 components.

@@

    To improve the convergence rate of the optimization process we employ the following optimization strategy: the optimization strategy begins with the optimizer viewing $\log{\mathbf{x}}$, and the first step involves running a rough estimate using a uniform prior. This step is optional, but if chosen, the next step involves optimizing the parameters associated with the systematic error described in Eq.~\ref{eq:model-npred-calibration}. Following this, a full optimization is performed using the patch prior, which can be computationally expensive. The optimization algorithm used is ADAM, a stochastic gradient descent method, without decay. Finally, there is an optional step of learning rate scheduling, although so far, a fixed step size value of $\mathrm{lr}=10^{-3}$ has been used.

@@

    \section{Experiments}
    \subsection{Test Datasets}
    \label{subsec:test-datasets}
    To evaluate the reconstruction performance of \jolideco and compare it against other methods we defined a set of test source scenarios. They consist of simplified elementary morphological structures resembling those found in astronomical images. We have devised four distinct arrangements of combinations of point and extended sources of different shapes. The scenarios are:

>>>>

\newcommand{\vlk}[1]{{\color{blue} [VLK: #1]}}
\newcommand{\as}[1]{{\color{purple} [AS: #1]}}
% Begin!
\begin{document}

@@

% Author list
    \author[0000-0003-4568-7005]{Axel Donath}
    \author[0000-0002-0905-7375]{Aneta Siemiginowska}
    \author[0000-0002-3869-7996]{Vinay Kashyap} 
    \author[0000-0000-0000-0000]{David van Dyk}
    \author[0000-0003-4428-7835]{Douglas Burke}

@@

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is often called \enquote{deblurring} or \enquote{deconvolution}, as it aims to invert the effect of the data being degraded by a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is categorised as \enquote{inverse problems}. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as \enquote{unblind deconvolution}. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant \as{I would say stars instead of AGN - in optical and even in X-rays telescopes AGN are resolved, so not point sources} activate galactic nuclei or a model of the PSF can obtained from simulation of the telescope or instrument.

\as{You may want to add a short para or just a sentance saying that X-ray and gamma-ray images are the Poisson type images containing low number of counts in each image pixel etc.. - this is to link to the next subsection which is specific to Poisson noise?}

    \subsection{Related Work}
    In literature there have been many efforts to solve the unblind deconvolution problem for images affected by Poisson noise. Most methods rely fundamentally on modeling the image generating process as a linear process. Given a known PSF a prediction of the discrete measured image can be obtained via \enquote{forward folding} and be compared against an observation assuming Poisson statistics per pixel. However, the methods differ in their additional prior or regularisation assumptions as well as their approach to optimization or sampling of the underlying likelihood function. In the following section we give a short overview over those methods.

@@

    However, in practice the RL method was found to have multiple, partly related limitations: first there is no objective criterion for when to stop the optimization process. Usually the iteration is stopped when the reconstructed image seems plausible to the eye. Second, when the number of iterations is too high, the method shows a tendency to decompose regions of constant brightness into a set of bright emission peaks (\enquote{speckles}). This problem was for example described and studied by \cite{Reeves1995} and \cite{Fish1995}. Lastly, the RL method only provided a \enquote{point estimate} and no associated estimate of the uncertainties of the reconstruction, which is desirable especially in the Poisson domain, for example, to measure the significance of specific features found in the reconstructed image.

    To address the issue with RL decomposing extended emission structures into groups of point sources, multiple additions to the RL method were proposed. Most importantly this lead to the introduction of regularised RL methods, such as use of total variation \citep{Dey2006} or sparse representations based on \enquote{wavelets} or \enquote{curvelets} \citep{Starck2003}. The challenge is rooted in two fundamentally contradicting goals: a deconvolution process that favors point sources will decompose extend structures present in the image, while introducing local correlations between neighbouring pixels will lead to less sharp reconstruction of point sources.

    One possible approach to overcome this challenge has been proposed by \cite{Esch2004} and \cite{Connors2011} as the \enquote{Low Counts Image Reconstruction and Analysis} (LIRA) method. Their key idea was to introduce a prior assumption that favors smoothness on multiple spatial scales of the image. Instead of a trying to find a single point estimate for the reconstructed image they also introduced the sampling from the posterior distribution, which allowed them to quantify the associated uncertainties. However, the prior distribution is based on the generic assumption of smoothness and does not encode any \enquote{real world} or physical information on the structures present in the image.

    An interesting modification of the standard RL method has been proposed by \cite{Ingaramo2014}. They showed that RL can be effective in creating a merged image from multiple observations. They also observed that RL would take the best visual properties out of each image and merged them to create a more accurate estimate of the `ground truth'.

    \cite{Selig2015} and \cite{Pumpe2018} approached the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they were able to deconvolve an astrophysical observation as well as decomposing it into point-like and diffuse emission components. While the method has worked well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it seemed to fail for smaller extended emission structures such as morphological details of extended sources.

    Other interesting approaches to the problem of deconvolution can of course be found in the field of computer vision. For example \cite{Zoran2011} build on very similar assumption of local correlation for natural images. However, instead of assuming a fixed local correlation structure, such as smoothness defined by a Gaussian with a correlation size, they proposed to learn the local correlation structure on millions of patches extracted from a set of reference images. For this they proposed to split training images from representative data sets into small patches and learn a  Gaussian Mixture Model (GMM) modeling the distribution of extracted patches. To use the GMM in reconstruction they introduced the concept of expected patch likelihood (EPLL). They showed that the patch based GMM prior led to much improved image reconstructions for multiple inverse problems such as denoising, deblurring and inpainting. \cite{Bouman2016} later adapted the patch prior successfully for the reconstruction of sub-millimeter radio data from the Event Horizon Telescope (EHT).

@@

    \subsection{Some Remarks on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the standard for deblurring and super-resolution of natural images. However, their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical \xray and \gammaray data.

    Firstly, astronomical \xray and \gammaray data are rare, expensive to obtain and there is no ground truth, because we can only observe the sky with our given instruments. To circumvent this issue one could rely on transfer learning and simulated data. For example radio or optical images could be forward folded with a known PSF model and be degraded with Poisson noise to generate training data. However, astronomical images still have a large dynamic range and show a high diversity in morphological structure, which would still require a large amount of training data. However, this approach has been successfully applied for specific classes of sources, such as images of Galaxies in the Sloan Digital Sky Survey (SDSS) \citep{Schawinski2017} or as a post processing step of results obtained with regularised RL \citep{Akhaury2022}. Recently, \cite{Sweere2022} used images from
    the large scale structure simulations with promising results.

@@

    One last important challenge for deep learning based methods is obtaining uncertainty estimates for the reconstructed image from CNNs. In the limit of Poisson noise it is desirable to trace the full a-posteriori likelihood, which is not possible for most network architectures, with the exception of, for example, normalising flows. This problem could be addressed with Monte Carlo (MC) based methods, such as bootstrapping, which further increases the demand on training data and computational resources. For these reasons we decided chose a statistical modelling approach to approach the de-convolution problem in the presence of Poisson noise. However, we will include elements from ML to represent the prior information on images.

@@

    \begin{itemize}
        \item Different observations with one instrument or telescope at different times and observation conditions. For example, multiple observations of \chandra of the same astrophysical object with different offset angles and exposure times;
        \item Observations from different telescopes, which operate in the same wavelength range. For example, a \chandra and \xmm observation of the same region in the sky;
        \item A single observation with one telescope with different data quality categories and different associated instrument response functions, such as event classes for \fermi.
    \end{itemize}

    Or even an arbitrary combination of the possibilities listed above. It could also be the combination of observations in different energy ranges, however in this case the assumption of the morphology not changing with energy would apply, which is rarely fulfilled for the astronomical data.

@@

    Where $\mathbf{\lambda_j}$ depends on $\mathbf{x}$ as defined by Equation~\ref{eq:model}. By taking the logarithm and dropping the constant terms one can transform the product into a sum over pixels, which is also often called the \textit{Cash} \citep{Cash1979} fit statistics:
    %
    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{D_j} | \boldsymbol{\lambda_j} \right) = \frac{1}{N}\sum_{i=1}^N \lambda_{j,i} - D_{j, i} \log{\lambda_{j,i}}
    \end{equation}

    To account for the possibility of different data sets having a different number of pixels we divide by $N$. By summing over all observations we finally get the joint log-likelihood of measuring a set of counts images $\mathbf{D}$ given an unknown flux image $\mathbf{x}$:
    %
    \begin{equation}
        \label{eq:joint}
        \mathcal{L}\left( \mathbf{D} | \mathbf{x} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D_j} | \mathbf{x} \right)
    \end{equation}

%\as{in Eq.4 - should there will be $\lambda$ not x in the sum term?, or maybe this is correct, but the formula in 3 is not? I'm confused.}

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate of $\hat{\mathbf{x}}$ directly, using some optimization procedure. However, in general this represents an \textit{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the existence of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

@@

    \subsection{Patch Based Image Prior}
    In general it is difficult to capture global image statistics in a prior likelihood taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the number of pixels squared and thus quickly becomes computationally untraceable. However, astronomical images typically only show local correlations between pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light (structure in the CMB might be an exception here). This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the idea of a patch based image prior approach as introduced by \cite{Zoran2011}. Given a pre-trained Gaussian Mixture Model (GMM) for the distribution of pixel values in patches, we can evaluate a patch based prior for the reconstructed image $\mathbf{x}$ and approximate the \aposteriori likelihood.

    One convenient property of the G MM is the possibility to evaluate its likelihood in closed form. For a given single patch $\mathbf{x}_n$ it is given by:
    %
    \begin{equation}
        \label{eq:gmm}
        P_{GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k \mathcal{N}(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2)
    \end{equation}

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the most likely GMM component per patch, becomes a discrete hyper-parameter in the model that is optimized along with the image $\mathbf{x}$. The bottom row of Figure~\ref{fig:patches} shows an example of a patch and the same patch represented in the \enquote{Eigenbasis} of the covariance matrices of its three most likely GMM components. Visually, the representation of the patch in the Eigenbasis of the covariance results in a denoised version of the patch.

@@

    \cite{Zoran2011} found that the reconstruction quality could be improved by working with a grid of overlapping patches. Such a grid is illustrated in Figure~\ref{fig:patches}. However, using a grid of overlapping patches makes the interpretation of the prior as likelihood more difficult, because pixels at the corners and edges of patches are counted multiple times. For this we introduce a weighting mask when evaluating the likelihood of the GMM. This mask weights the pixels of the patch by their overlap with the neighboring patch. Overlapping edges are assigned a weight of 1/2, while overlapping corners are assigned a weight of 1/4, when computing the likelihood for a patch as given by  Equation~\ref{eq:gmm}. Apart from this we keep the process of choosing the most likely component unmodified. However, we note that in this case we neglected the correlation between neighbouring patches. As the choice of the GMM component per patch is discrete, a simultaneous gradient based optimization of all patches is not easily possible, because we cannot compute gradients across GMM components. This would require a continuous representation or interpolation between GMM components.

@@

    To normalise the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to be present at many levels of intensity. While this could be handled by introducing a lot more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we propose the following modified normalisation for the patches instead:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \tilde{\mathbf{x}}_n = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}
    \end{equation}

@@

    \subsubsection{Cycle Spinning}
    The evaluation of the patch prior divides the reconstructed image into a fixed grid of overlapping patches. To avoid artifacts in the reconstruction due to the fixed position of the grid, we use the method of \enquote{cycle spinning}. Before evaluating the patch prior we randomly shift the image $\mathbf{x}$ by a given number of pixels in x and y direction. Were the shift is smaller than half the size of a patch. The same method has, for example, been used in \cite{Esch2004}. We also experimented with cycle spinning by a sub-pixel amount, mostly for the reconstruction of point sources. However, we later introduced the possibility to reconstruct an upsampled version of $\mathbf{x}$, which made the sub-pixel cycle-spinning mostly redundant.

@@

    \subsection{Systematic Errors of Predicted Counts}
    Each observation $\mathbf{D_j}$ comes with associated systematic errors. This includes for example the pointing accuracy of the telescopes: when repeatedly pointed to the same object or when using different types of telescopes for observation, the absolute position of an astronomical object might vary in the image between different observations. The process of correcting for these inaccuracies is called \textit{astrometry} and a standard procedure in for example \chandra data analysis. A second example of systematic errors is the variation of background emission between multiple observations.

@@

    \subsection{Estimation of Statistical Errors}
    Estimating the global covariance matrix for image based data is computationally not feasible, because the size of the matrix scales with the squared number of pixels in the image. However, there are a few ways to estimate the statistical uncertainties associated with the flux image. We propose three methods for this:

    \begin{itemize}
    \item The first method is bootstrapping of observations. Given that multiple observations of the same region of the sky are available, \jolideco can be run repeatedly on a random subset of those. This results in a set of reconstructed images which are sampled from the posterior distribution. Depending on the size of the images the associated computational effort can be very large as it requires running \jolideco many times to achieve sufficient statistics.

    \item If only one or few observations are available we propose bootstrapping of the events used to compute the counts image instead. This basically corresponds to splitting a single observation into multiple ones by splitting the single exposure into multiple time intervals. This way one can apply the first proposed method again.

@@

\as{There is no mention of stochastic gradient decent in the description of the method above, so this is the first time you talk about it... maybe it needs to be included when you describe solving the GMM model?}

@@

    \subsection{Jolideco GMM Library}
    \label{ssec:jolideco-gmm-library}
    In general the strength of the GMM prior is its high \textit{expressiveness}. This means the prior contains a large variety of structural shapes and has a lot of freedom to adapt to the data in the optimization process. On radio data \cite{Bouman2016} found their results to be mostly independent of the choice of data the GMM prior was learned on. However, in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of the data the GMM was learned on. The choice of the prior becomes more important and to avoid introducing a strong bias it should be adapted to the analysis scenario. For this reason and for purposes of testing we learned and provide a selection of GMMs (a \textit{Library}) to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

@@

    \subsubsection{GLEAM Radio Data}
    \label{sssec:gleam-radio-data}
    As an alternative we consider a prior learned from astronomical radio data. For this we relied on high signal to noise radio data from the \textit{GLEAM} survey \cite{HurleyWalker2022}. We downloaded the data using the Gleam VO Client\footnote{\url{https://github.com/ICRAR/gleamvo-client}}. We restricted the choice of data to the Galactic Plane between $b=\pm10\deg$ and $\ell = \pm180\deg$ to avoid a bias towards extra-galactic point sources and represent structures of Galactic sources with rich morphology, such as supernova remnants, pulsar winds etc. The dataset still includes Galactic as well as extra-galactic point sources in the background. In total we extracted $\approx 6 \times 10^5$ patches and trained a GMM with 128 components. We used the patch normalisation described in Section~\ref{ssec:patch-norm}.

@@

    \subsubsection{Chandra SNRs}
    \label{sssec:chandra-snrs} Because of the limited statistics we did not use any oversampling but kept the pixel. Astrophysical processes forming the morphological structures are typically self-similar. Structures appear on multiple angular scales. This means the prior could even be learned from the data itself if sufficiently free of noise. For analysis of Galactic sources with complex morphology and typically radially symmetric features such as supernova remnants (SNRs) we learned a prior from images of the bright \chandra SNRs \textit{Cas A} and \textit{G292.0+01.8}. In total we extracted $\approx 10^6$ patches and learned a GMM with 128 components.

@@

    To improve the convergence rate of the optimization process we employ the following optimization strategy: the optimization strategy begins with the optimizer viewing $\log{\mathbf{x}}$, and the first step involves running a rough estimate using a uniform prior. This step is optional, but if chosen, the next step involves optimizing the parameters associated with the systematic error described in Eq.~\ref{eq:model-npred-calibration}. Following this, a full optimization is performed using the patch prior, which can be computationally expensive. The optimization algorithm used is ADAM \citep{Kingma2014}, a stochastic gradient descent method, without decay and a fixed step size value of $\mathrm{lr}=10^{-3}$ has been used.

@@

    \section{Experiments}
    \subsection{Test Datasets}
    \label{subsec:test-datasets}

    \vlk{We have devised several test patterns that are designed to exercise and stress test deconvolution algorithms for low counts images.  The patters consist of both point and extended sources, with strengths characteristic of typical observations, arranged with different spacings between the sources, and include point sources superposed on extended sources, as well as different background levels.  The goal of the patterns is not to set up or devise a ``perfect'' deconvolution algorithm, but to help assess their relative strengths and limitations.  The source patterns are described in Appendix~\\ref{sec:appA}.

    We adopt two distinct scenarios, one which mimics \textit{Chandra} observations with a sharp PSF (a Gaussian with $\sigma=2$~pix) and low background (at 0.01, 0.1, and 1 ct~pix$^{-1}$) but lower effective area (nominally with an exposure of 1 unit), and the other which mimics XMM-\textit{Newton} observations with a broader PSF ($\sigma=3$~pix), higher background (at 0.1, 1, and 10 ct~pix$^{-1}$), and higher effective area (imitated with an exposure of 5 units).
    }

    % To evaluate the reconstruction performance of \jolideco and compare it against other methods we defined a set of test scenarios. They consist of simplified elementary morphological structures resembling those found in astronomical images. We have devised four distinct arrangements of combinations of point and extended sources of different shapes. The scenarios are:
